\chapter{Introduction}

\section{Introduction}

This document is not a tutorial and assumes the reader has some knowledge of programming in general and BASIC in particular. \\

If you want to learn programming, I would advise you try learning using one of the numerous courses available. The course in the Vic20 manual is highly regarded and can be used with an emulator such as "Vice", and much of the knowledge will transfer directly.\\

F256's SuperBASIC is a modernised BASIC interpreter for the 65C02 processor. It currently occupies 4 pages (32k) of Flash mapped into 2 pages (16k) of the 65C02 memory space. At present, this is from $8000-$BFFF.\\

Although line numbers are still used for editing, they are no longer the primary mechanism for control flow. Cross-development without line numbers is fully supported, with examples available in the \path{games} directory of the SuperBASIC GitHub repository.\footnote{\url{https://github.com/FoenixRetro/f256-superbasic}}\\

GOTO, GOSUB, and RETURN remain available for compatibility with legacy code, but their use is discouraged. For greater clarity and maintainability, it is recommended to use procedures and structured constructs such as FOR/NEXT, WHILE, REPEAT, and IF/ELSE/ENDIF.\\

\section{Storage}

Programs are stored in ASCII format, so in cross-development, any editor can be used. LOAD, VERIFY, and SAVE read and write files in this format. Internally, the format is quite different.

\section{Memory usage}

Memory usage is split into 2 parts. \\

The main space from \$2000-\$7FFF contains the tokenised BASIC code. Keywords such as REPEAT are replaced by a single byte, or for less common options, by two bytes.\\ 

Identifiers are replaced by a reference into the identifier table from \$1000-\$1FFF. The first part of this table is a list of identifiers along with the current value.\\

Arrays and allocated memory (using ALLOC) follow that.\\

Finally, string memory occupies the top of this memory area and works down.\\

This should be entirely transparent to the developer.

\section{Memory usage elsewhere}

SuperBASIC uses memory locations outside the normal 6502 address space as well. \\

If you use a bitmap, it will be placed at \$10000 in physical space and occupy 320x240 bytes.\\

If you use sprites, they are loaded to \$30000 in physical space, and the size depends on how many you have.\\

If you use tiles, the tile map is stored at \$24000 and the tile image data at \$26000. \\

Finally, if you cross-develop, the memory location from \$28000 onwards is used to store the BASIC code you have uploaded.\\

If you do not use any of these features directly (you can set up your own bitmaps and sprites yourself, and enter programs through the keyboard, saving to the SD Card or IEC drive) then the memory is all yours.
